Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-11-13 text/sections/00-abstract.tex
Adjustments.

US public universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically cut higher education funding despite rising enrolment.
This paper asks whether the changes the faculty composition at US public universities are caused by the fall in state funding, using a shift-share approach to instrument for state funding.
I find that universities employ $4.4$\% more lecturers per student in response to a fall of $10$\% per student funding, and correspondingly decrease their reliance on tenure-track faculty by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
%Dynamic estimates show that these effects linger multiple years after the initial funding cut.
In a secondary analysis, using data on all faculty at Illinois public universities, I find that incumbent professors' salaries, promotion rate, and quit rate are all not affected by funding cuts, implying that faculty composition change arose by limiting the hiring of new tenure-track/tenured professors at public universities.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

@@

\vspace{0.05cm}
\noindent
\textbf{JEL Codes:} H72, I23, J45

>>>>

US public universities have systematically substituted away from tenure-track/tenured professors towards lecturers over the last thirty years, while states have systematically cut public funding for higher education.
This paper shows that falling state funding caused changes in faculty composition, using a shift-share approach to instrument for state funding cuts.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
In a secondary analysis, using data on all faculty at Illinois public universities, incumbent professors' salaries, promotion rate, and quit rate are not affected by funding cuts, implying that composition change arose by limiting hiring of new tenure-track/tenured professors.
These results show institutional effects of stagnating public funding for education, exhibiting one possible mechanism for deteriorating student outcomes over this same time period.

@@

\vspace{0.05cm}
\noindent
\textbf{JEL Codes:} H72, I22, J45

### 2023-11-13 text/sections/01-introduction.tex
Adjustments.

Public universities have experienced a stagnation in state funding over the last three decades, despite a dramatic increase in enrolments.
At the same time, public unviersities have increasingly subsituted away from tenure-track faculty towards lecturers, who often work on contingenet and lower-salaried contracts.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether or how much this funding decline has impacted faculty.
In this paper, I show there is a causal relation between the trends of declining state funding for higher education and increasing reliance on lecturers.
In a secondary analysis of all public university faculty in the state of Illinois, I show  shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates.
This implies that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors.
Thus stagnating state funding for public universities has implications for the wider structure of US higher education and research.

It is a priori unclear how faculty may be affected, as there are various margins for university to respond to funding cuts.
A university's level of resources affect the ability that departments can hire new professors, retain their current faculty, or provide retirement incentives, so it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
However, over the same time period, higher education has undergone dramatic changes in the rate of selectivity thanks to greater competition \citep{hoxby2009changing}.
These changes may have lead public universities to resist relying on lecturers, and focus more on raising revenue from other sources, such as federal or private funds.

I beging the paper by documenting the dramatic decline in state funding for public universities over recent decades, and drastic changes in faculty composition over the same time frame.
While US universities are widely considered the highest performing in the world, there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 of state funding per student in 1990, and only \$8,300 in 2021.
This fall is driven by a stagnation in the absolute level of state funding for higher education in most states, combined with a large increase of 46\% in student enrolment at public universities over the same time period.
Similarly, the number of lecturers per student over doubled, while the number of tenure-track/tenured professors per students fell by $-23$\% from 1990 to 2021.

@@

In both analyses, I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts how much a university relies on state funding as a percent of revenues in a base period with the entire state's total funding per higher education student for each year, exploiting changes in funding and how much each university historically relied on state funding for higher education.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method thanks to the presence of time-series confounding between the funding shock and later years' level of state funding.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is non-contemporaneous.
    %Previous work ignored the potential time-series confoduning, and estimated linear instrumental variables.
}

@@

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding has examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, there is applied theory work that models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, several empirical papers measure how universities were affected by endowment shocks \citep{brown2014endowment}, and the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure effects of state funding and changes in university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.

>>>>

Public universities have experienced a stagnation in state funding over the last three decades, despite a dramatic increase in enrolments.
At the same time, public unviersities have increasingly subsituted away from tenure-track faculty towards lecturers, who often work on contingenet and lower-salaried contracts.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether this funding decline has impacted faculty.
In this paper, I show there is a causal relation between the trends of declining state funding for higher education and increasing reliance on lecturers.
In a secondary analysis of all public university faculty in the state of Illinois, I show  shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates.
This implies that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors.
Thus stagnating state funding for public universities has implications for the wider landscape of US higher education and research.

State funding for public universities has dramatically declined over recent decades, and the composition of faculty at these universities changed drastically over the same time frame.
This fall is driven by a stagnation in the absolute level of state funding for higher education in most states, combined with a large increase of 46\% in student enrolment at public universities over the same time period.
Similarly, the number of lecturers per student over doubled, while the number of tenure-track/tenured professors per students fell by $-23$\% from 1990 to 2021.
%While US universities are widely considered the highest performing in the world, there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities rely on their state governments for funding, and receive over a quarter of their total revenue from state governments.
An average public university received around \$11,600 of state funding per student in 1990, and only \$8,300 in 2021.

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
University administrations may directly cut salaries or lay-off incumbent staff; in 2015, the University of Illinois furloughed faculty around a budget crisis \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, some faculty positions may grow at universities in response to funding cuts.
Public universities do not restrict enrolment in response to funding cuts, so someone must teach the next year of students.
Lecturers can fill in teaching duties, at a lower cost for the universities than tenure-track professors, with less employment protections (e.g., no tenure).
However, over the same time period, higher education has undergone dramatic changes in the rate of selectivity thanks to greater competition \citep{hoxby2009changing}.
This selectivity pressure may temper universities from relying on lecturers, and instead lead them to fill in the funding gap by raising revenue from other sources, such as private funds or raising tuition.

@@

In both analyses, I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts how much a university relies on state funding as a percent of revenues in a base period with the entire state's total funding per higher education student for each year, exploiting changes in funding and how much each university historically relied on state funding for higher education.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method thanks to the presence of time-series confounding between the funding shock and later years' level of state funding.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is non-contemporaneous.
}

@@

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding has examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, previous applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, several empirical papers measure how universities were effected by endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.

### 2023-11-13 text/sections/02-data.tex
Newest revision.
Adjustments.

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information including finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment and faculty counts come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

>>>>

\subsection{Data Description}
The data used in this paper come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information including finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1990 refers to the academic year that ran August 1989 to July 1990.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment and faculty counts come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

### 2023-11-13 text/sections/06-discussion.tex
Adjustments.

These effects are persistent: public higher education in the US was not exposed to one large instance of funding cuts, but systematic funding short-falls across the last three decades.
The long-term estimates (i.e., \autoref{fig:count-lp}) show that state funding cuts have effects on universities for multiple years after the initial funding cut, and these effects are compounded by the fact that public universities received persistent funding cuts for so many years.
The end result is large changes in faculty composition, the long-run trend in stagnating state funding leading to a long-run substitution towards lecturers at public universities.

>>>>

These effects are persistent: public higher education in the US was not exposed to one large instance of funding cuts, but systematic funding short-falls across the last three decades.
The long-term estimates (\autoref{fig:count-lp}) show that state funding cuts have effects on universities for multiple years after the initial funding cut, and these effects are compounded by the fact that public universities received persistent funding cuts for so many years.
The end result is large changes in faculty composition, the long-run trend in stagnating state funding leading to a long-run substitution towards lecturers at public universities.

### 2023-11-13 text/sections/09-figures-tables.tex
Adjustments.

\newpage
\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for Effect of State Funding on Faculty Count per Student at Universities, by Professor Group.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturers.}
        \includegraphics[width=\textwidth]{figures/lecturer-count-lp.png}
        \label{fig:lecturer-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant Professors.}
        \includegraphics[width=\textwidth]{figures/assistant-count-lp.png}
        \label{fig:assistant-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full Professors.}
        \includegraphics[width=\textwidth]{figures/full-count-lp.png}
        \label{fig:full-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All Professors.}
        \includegraphics[width=\textwidth]{figures/all-count-lp.png}
        \label{fig:all-count-lp}
    \end{subfigure}
    \label{fig:count-lp}
    \justify
    \footnotesize
    \textbf{Note}:
    These figures show the local projections estimates of regression specification \eqref{eqn:secondstage}, with the funding shock as an instrument for state funding
    The unit of analysis is the university, and uses IPEDS data.
    The coefficient estimate is effect of state funding ($X_{i,t}$) on faculty count per student ($Y_{i,t}$), using the funding shock instrument ($Z_{i,t}$), while accounting for auto-correlation between different time periods --- i.e., between $X_{i,t}, X_{i,t-1}$ and $Y_{i,t}, Y_{i,t-1}$.
    These results use a $\log-\log$ specification, so the estimates are for the elasticity of professor count per student in a year $t+k$ with respect to state funding in year $t$, where years $k = 0, \hdots, 10$ are on the $x$-axis. 
    Standard errors are clustered at the state-year level.
\end{figure}

@@

\newpage
\begin{table}[h!]
    \singlespacing
    \centering
    \caption{First Stage Estimates, for State Funding by Funding Shock in IPEDS Data.}
    \textbf{Panel A: units in \$ per student}

@@

\newpage
\begin{table}[H]
    \singlespacing
    \centering
    \caption{Shift-Share Instrument Balance Test, in IPEDS 1990--2017.}

    \textbf{Panel A: units in \$}

    \makebox[\textwidth][c]{\input{tables/firststage-balance-count.tex}}

    \makebox[\textwidth][c]{\input{tables/firststage-balance-log.tex}}

    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the OLS estimates of regression specification \eqref{eqn:firststage}, showing the effect of the funding shock on other outcomes to gauge balance of the instrument with respect to university characteristics
    Enrolment is measured in thousands of students.
    Each observation is a public universit--year, in the IPEDS data.
    Panel A shows the effect of a funding shock of \$$-1,000$ (per student in the state) on the outcome --- i.e.,
    \$$-1,000$ funding shock per student in the state is associated with 517 more students (as enrolment is measured in thousands) in column 1.
    Panel B shows the effect of a $-10$\% change funding shock per student in the state on $10$\% change in the outcomes --- i.e.,
    $-10$\% funding shock per student in the state is associated with an increase in 1.07\% more students at the university according to preferred column 1.
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included where noted.
    \label{tab:firststage-balance}
\end{table}

>>>>

\newpage
\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for Effect of State Funding on Faculty Count per Student at Universities, by Professor Group.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturers.}
        \includegraphics[width=\textwidth]{figures/lecturer-count-lp.png}
        \label{fig:lecturer-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant Professors.}
        \includegraphics[width=\textwidth]{figures/assistant-count-lp.png}
        \label{fig:assistant-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full Professors.}
        \includegraphics[width=\textwidth]{figures/full-count-lp.png}
        \label{fig:full-count-lp}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All Professors.}
        \includegraphics[width=\textwidth]{figures/all-count-lp.png}
        \label{fig:all-count-lp}
    \end{subfigure}
    \label{fig:count-lp}
    \justify
    \footnotesize
    \textbf{Note}:
    These figures show the local projections estimates of regression specification \eqref{eqn:secondstage}, with the funding shock as an instrument for state funding.
    The unit of analysis is the university, and uses IPEDS data.
    The coefficient estimate is effect of state funding ($X_{i,t}$) on faculty count per student ($Y_{i,t}$), using the funding shock instrument ($Z_{i,t}$), while accounting for auto-correlation between different time periods --- i.e., between $X_{i,t}, X_{i,t-1}$ and $Y_{i,t}, Y_{i,t-1}$.
    These results use a $\log-\log$ specification, so the estimates are for the elasticity of professor count per student in a year $t+k$ with respect to state funding in year $t$, where years $k = 0, \hdots, 10$ are on the $x$-axis. 
    Standard errors are clustered at the state-year level.
\end{figure}

@@

\newpage
\begin{table}[H]
    \singlespacing
    \centering
    \caption{First Stage Estimates, for State Funding by Funding Shock in IPEDS Data.}
    \textbf{Panel A: units in \$ per student}

@@

### 2023-11-13 text/sections/10-appendix.tex
Adjustments.

\subsection{IPEDS First Stage}

@@

\newpage
\subsection{Additional Results, Faculty Hiring}
\label{sec:appendix-hiring}

@@

\newpage
\subsection{Additional Results, Rates of Substitution}
\label{sec:appendix-substitution}

>>>>

\subsection{IPEDS First Stage}
\begin{table}[H]
    \singlespacing
    \centering
    \caption{Shift-Share Instrument Balance Test, in IPEDS 1990--2017.}

    \textbf{Panel A: units in \$}

    \makebox[\textwidth][c]{\input{tables/firststage-balance-count.tex}}

    \makebox[\textwidth][c]{\input{tables/firststage-balance-log.tex}}

    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the OLS estimates of regression specification \autoref{eqn:firststage}, showing the effect of the funding shock on other outcomes to gauge balance of the instrument with respect to university characteristics
    Enrolment is measured in thousands of students.
    Each observation is a public universit--year, in the IPEDS data.
    Panel A shows the effect of a funding shock of \$$-1,000$ (per student in the state) on the outcome --- i.e.,
    \$$-1,000$ funding shock per student in the state is associated with 517 more students (as enrolment is measured in thousands) in column 1.
    Panel B shows the effect of a $-10$\% change funding shock per student in the state on $10$\% change in the outcomes --- i.e.,
    $-10$\% funding shock per student in the state is associated with an increase in 1.07\% more students at the university according to preferred column 1.
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included where noted.
    \label{tab:firststage-balance}
\end{table}

@@

\newpage
\subsection{Faculty Hiring}
\label{sec:appendix-hiring}

@@

\newpage
\subsection{Rates of Substitution}
\label{sec:appendix-substitution}
